Temporal Evolution of Erosion in Pore-Networks:\\ From Homogenization to Instability

### 2022-03-06 prl_paper.tex
Updates from Overleaf

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%%
\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose,
%preprint,
%showpacs,preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}

@@

\usepackage{xcolor}
\newcommand{\DP}[1]{\noindent \color{cyan} (DP: #1)\normalcolor}
\newcommand{\AZ}[1]{\noindent \color{purple} (AZ: #1)\normalcolor}
\newcommand{\add}[1]{\noindent \color{blue} #1 \normalcolor}
\newcommand{\del}[1]{\noindent \color{red} \st{#1}\normalcolor}
\newcommand{\AM}[1]{\noindent \color{magenta} (AA: #1)\normalcolor}
\def\red{\textcolor{red}}
%
\begin{document}
%
\preprint{APS/123-QED}

\title{Temporal Evolution of flow in Pore-Networks:\\ From Homogenization to Instability} 
% Force line breaks with \\
% \thanks{A footnote to the article title}%

\author{Ahmad Zareei}
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break automatically or can be forced with \\
\author{Deng Pan}%
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break automatically or can be forced with
\author{Ariel Amir}%
\email{arielamir@seas.harvard.edu}
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break  automatically or can be forced with
% need to figure out how to reference dagger

@@

\begin{abstract}
We study the dynamics of flow-networks in porous media using a pore-network model. First, we consider a class of erosion dynamics assuming a constitutive law depending on flow rate, local velocities, or shear stress at the walls. We show that depending on the erosion law, the flow may become uniform and homogenize, stay as it is, or become unstable and develop channels. By defining an order parameter capturing different behaviors we show that a phase transition occurs depending on the erosion dynamics. Using a simple model, we identify quantitative criteria to distinguish these regimes. Lastly, we show that pores clogging at the initial stages show analogous behaviors depending on clogging dynamics. % , however, due to the pore throat blockages and changes in the network connections the evolution diverges from its initial trend.

@@

Fluid flow through a porous medium undergoing a dynamical change in its network of micro-structure is a challenging problem and has many environmental and industrial applications~\cite{fraggedakis2015flow,sahimi2011flow,schlesinger1999carbon,herzig1970flow,tien1979advances,jaisi2008transport,carrel2018biofilms,seymour2004anomalous,duduta2011semi,sun2019hierarchical,marbach2016pruning,tero2010rules,alim2013random,heaton2010growth}.
% , such as oil recovery~\cite{fraggedakis2015flow,sahimi2011flow}, CO$_2$ sequestration~\cite{schlesinger1999carbon}, water filtration/separation~\cite{herzig1970flow,tien1979advances,jaisi2008transport,carrel2018biofilms,seymour2004anomalous}, energy storage~\cite{duduta2011semi,sun2019hierarchical}, or biological applications \cite{marbach2016pruning,tero2010rules,alim2013random,heaton2010growth}. 
The disordered pore structure of a porous medium results in a heterogeneously distributed fluid flow between the pores.
% Fluid flow in a porous medium is often heterogeneously distributed between the pores as a result of a highly disordered pore structure.
Due to the fluid flow, the pore structure can further change dynamically either through the erosion of pore boundary walls or deposition/sedimentation of material on the boundary of the pores. Such heterogeneous and dynamical changes of the solid structure affect the pore-level fluid flow which in turn affects the dynamical changes to the pore structure. This feedback mechanism along with the initial heterogeneous fluid flow complicates the dynamical process and makes it difficult to understand and predict the porous media behavior. Nonetheless, an understanding of the dynamical change is essential to improve any of the porous media applications where the pore network changes over time, applications such as groundwater remediation and precipitation of minerals in rocks~\cite{rad2013pore}, biofilm growth in water filtration, and protective filters \cite{herzig1970flow,tien1979advances,jaisi2008transport,carrel2018biofilms,seymour2004anomalous}, as well as enhanced oil recovery with polymer flooding \cite{lake2014fundamentals,parsa2020origin}, or water-driven erosion \cite{schorghofer2004spontaneous,mahadevan2012flow}.

% to the pore structure affects local fluid flow as well as bulk properties of the medium. 
% Understanding flow behavior is important to improve any of the the applications where the pore structure changes dynamically either by erosion of the boundary wall or deposition/sedimentation of material on the boundary. Changes in the micro-structure and solid boundaries affect local fluid flow as well as the bulk properties. The  erosion or deposition process in a porous media plays an essential role in many environmental and industrial applications such as groundwater remediation and precipitation of minerals in rocks, bio-film growth in water filtration and the design of protective filters,  as well as enhanced oil recovery with polymer flooding, water-driven erosion. The structures of such porous media are usually highly disordered with a heterogeneous distribution of pore-level flow velocities. 
%%

% Albeit there is an abundance of applications in understanding 
% Despite the importance of understanding flow evolution in porous structures exposed to erosion and deposition,it 
% \footnote{It can be shown that the rate of change in tube's radius in \cite{reddi2000permeability} assuming a dilute solution becomes proportional to local fluid velocity},

In light of the plethora of applications, it is surprising that the evolution of porous structures exposed to erosion and deposition has only been partially understood both theoretically and experimentally. In order to model erosion in porous materials, different models have been used where erosion is assumed to be locally proportional to different flow parameters. Some studies have focused on the erosion of the pore structure proportional to the shear stress at the walls \cite{jager2017channelization,ristroph2012sculpting,hacking1996shear,wan2004investigation}, or proportional to the power dissipated by the flow~\cite{steeb2007modeling,marot2012study,sibille2015internal}. 
More recently it has been shown that using a coarse-grained model with an erosion rate depending on the local pressure gradient, different branching instability can be reached in a porous material~\cite{derr2020flow,mahadevan2012flow}.  
Similarly, in clogging, deposition rate based on local fluid flux, velocity, or even at a constant rate independent of any flow parameters has been used~\cite{yang2017pore,reddi2000permeability,boccardo2014microscale}. In order to unify the different approaches, we use a general continuous model which allows us to study the effect of different erosion or clogging dynamics with a single constitutive law. Additionally, we use a network approach~\cite{alim2017local,fatt1956network, blunt2013pore,stoop2019disorder,bryant1993network} to study the flow dynamics in the porous media.  The network approach has the advantage of not requiring coarse-grained homogenization and is computationally simple allowing us to simulate larger domains.

% (which is the same as shear at the walls),
% to capture the flow dynamics

% Despite all of the different models, a consensus on the dynamics of erosion or clogging is missing.\red{focus more on the network model} The adventage versatile law network approach -> does not need homogenization or meanfield assumption

% Since the dynamics of erosion in a porous material is not well understood,

% A phenomenological discrete network-based model~\cite{alim2017local,fatt1956network, blunt2013pore,stoop2019disorder,bryant1993network} is used to model the hydrodynamically driven erosion or clogging in a porous material.
% In the network model, the pores are represented by spherical nodes connected with edges represented by cylindrical tubes.

The network of pores inside the solid structure is connected together through pore-throats that effectively show resistance to the fluid flow between the pores (Fig.~\ref{fig:fig1}a). In a network model, the pores/voids of the porous medium are represented by a two- or three-dimensional lattice of nodes. The edges between the nodes of the network represent the pore-throats between the pores and are modeled using cylindrical tubes characterized by their radius and length. 
Network-based models have successfully shown to capture key properties of fluid flow in a porous material such as the probability distribution of fluid flux~\cite{alim2017local}, the permeability scaling during clogging~\cite{shima2021}, or the first fluidized path in a porous structure~\cite{fraggedakis_chaparian_tammisola_2021}. 
% Using such network models, it has been shown that the probability distribution of fluid flux in a porous material is universal when the porous material is random enough. Furthermore, during the clogging process, the usage of Kozeny-Carman fails to correctly predict the porosity since all the length scales are not eroding at the same time (the erosion only happens at the corss sections and not along the length of a porous material). New relation based on the variation in porosity has been proposed that can successfully find the relation between porosity and permeability variation using such models which is verified experimentally.
The degradation of the solid skeleton (i.e., erosion) or deposition of material on the pore throats (i.e., clogging) in such networks is modeled by the change (increase or decrease) in the radius of the edges connecting the pores which ultimately translates in the changes in the flow resistance between the pores. Using a general local erosion law that can model different dynamics, we show a variety of behaviors are possible. The network can either move toward stability and homogenization, stay as it is with no significant change in network statistics, or become unstable and develop channels. Using numerical solutions with an erosion model on a random network we show the emergence of two behaviors and morphologies are possible through selective erosion and subsequent flow enhancement. Using a simplified model we capture the underlying physics and furthermore show a phase transition occurs depending on the parameters of the model. 
% Here, based on the porous network model, we link various erosion laws with different large-scale behaviors and show a variety of behaviors that can be observe
% In order to better understand the effect of different erosion dynamics in a porous based on local parameters, 
% network model has been successful in predicting the first open channeel \cite{fraggedakis_chaparian_tammisola_2021} and capturing the fluid flow distribution \cite{alim2017local}
% \cite{fatt1956network} 
% \cite{blunt2013pore}
% \cite{stoop2019disorder}
% Consequently, clogging effects are excluded from these assumptions. Thus, the eroded fines can be transported through the porous material within these statistically homogeneous distributed channels
% Albeit there is an abundance of applications, the evolution of structures exposed to erosion and deposition is in general not easily predictable and usually requires the use of numerical techniques.
% However, this process is difficult to model due to diverse processes that may arise, including particle advection through the pore space by fluid flow, adsorption and deposition onto the solid matrix, and erosion or resuspension.
% \begin{itemize}
% \item Importance of porous media
% \item importance of dynamics in porous media
% \item we study evolution with a network model
% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
% \section{Methods \& Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\begin{figure}[h]
    % \centering
    \includegraphics[width = 0.45\textwidth]{Figs/Fig1_pppp.eps}
    \caption{(a) Cross section of a porous media obtained using  computerized tomography (CT) scan of a sandstone sample (CT image taken from \cite{akanji2010finite}). The scale in the bottom left shows $1$mm. The network of pores and throats is highlighted in blue. In the network model, the pores are represented with nodes and the throats between pores are approximated with tubes connecting the pores together. (b) Schematic of a topologically random network (i); and a structured diamond grid network (ii). The edge diameters in both networks are randomly distributed. The inset figure shows the conservation of mass at each node. (c) The universal probability distribution function (PDF) of fluid flux for a topologically random (in blue) or diamond network (red) with a highly disordered random net including uniform (triangles), log-normal (circles), and truncated normal (plus) distributions. The distribution parameters are chosen such that the coefficient of variation (the ratio of the standard deviation to the mean) becomes $0.5$ which results in a highly disordered pore throat distribution. Generally, in a highly disordered random network, the PDF of normalized fluid flux becomes independent of the network and follows a universal distribution with an exponential tail. }\label{fig:fig1}
\end{figure}
%
% \AZ{Things are not topoligically random }
% \AZ{random end }
% \AZ{structured allows for mean field theory}
% \AZ{the edges are the same - exponential where it is coming from}
% \AZ{boundary between phases, where the mean fields hold}

  % Flow through pores and throats of a porous medium can be modeled using a pore network model \cite{alim2017local,fatt1956network, blunt2013pore,stoop2019disorder}. In a network model, the pores/voids of porous medium are represented by a two- or three-dimensional lattice of nodes. The edges between the nodes of the network represent the pore-throat between the pores and are modeled using cylindrical tubes characterized by their radius and length (see Fig. \ref{fig:fig1}(a)). The pore-network model is well suited for the simulation of low-Reynolds fluid flow and has proved to capture the flow dynamics of porous media \cite{alim2017local} and also the first fluidized path \cite{fraggedakis_chaparian_tammisola_2021}. We use a structured/unstructured network with randomly distributed nodes and edge length. 
% It has been shown that flow through such network correctly captures the exponential tail for distribution of fluid velocities through a porous medium.

\textit{Methods and Results}-- We consider low-Reynolds fluid flow through the porous network, i.e.,  $\rho u r/\mu \ll 1$ where $\rho$ is the fluid density and $\mu$ is the kinematic viscosity, $u$ and $r$ are characteristic fluid velocity and pore radius. In such conditions, the fluid flow in each tube follows the Poiseuille law
 $ p_i - p_j  = ({8 \mu l_{ij}}/{\pi r_{ij}^4}) q_{ij}$ where $p_i,p_j$ represents pressures at neighboring nodes $i$, $j$, and $r_{ij}, l_{ij}, q_{ij}$ are respectively the radius, length, and fluid flow rate at the edge connecting pores $i$ and $j$. The coefficient ${\pi r_{ij}^4}/{8 \mu l_{ij}}$ which relates the fluid flux to the pressure difference can be considered as the conductance $C_{ij}$ of the edge $ij$ in the pore-network. In addition to Poiseuille flow at the edges,  the total incoming flow at a node should be equal to the total outgoing flow in order to conserve mass. The conservation of mass at pore $i$ can be written as $\sum_{j\in \mathcal{N}(i)} q_{ij} =0$ where $\mathcal{N}(i)$ represents all the neighboring nodes of node $i$. First, we consider a topologically random network of nodes constructed using a uniform distribution of $N_x\times N_y$ nodes in a planar domain $\left[0,L_x\right]\times \left[0,L_y\right]$ where the nodes connectivity are obtained using Delaunay triangulation (see Fig. \ref{fig:fig1}b(i) for a part of such network). The radius of the edges is considered as independent and identically distributed random variables sampled from a probability distribution that can be either uniform, log-normal, or a truncated normal distribution (see supplementary material S1 for more details).  Assuming a pressure difference between the boundary nodes at the left and the right boundary, the pressure at the nodes, and the fluid flow rate at the tubes can be obtained by solving the corresponding set of linear equations. We find that for large enough randomness in the edge radii (i.e., $\text{std}(r_{ij})/\text{mean}(r_{ij})\geq 0.5$), the PDF is well described by a single exponential distribution as shown in Fig. \ref{fig:fig1}c. The exponential form of the PDF reflects the relatively small number of edges with extremely large fluid fluxes. The exponential distribution of fluid flux obtained here is similar to the earlier experimental or numerical measurements~\cite{datta2013spatial,shima2021,alim2017local}.  Next, we consider a structured diamond-grid of pores (Fig. \ref{fig:fig1}b(ii)) which significantly simplifies the geometrical complexity of the network. First, we find that the PDF of normalized fluid flux remains unchanged in such networks for various distributions and in fact matches with an unstructured random network (Fig. \ref{fig:fig1}c). The diamond grid allows us to analytically track the propagation of fluid into the network and calculate the distribution of fluid flux using a mean-field approximation (see supplementary material S3). Using a similar idea proposed for the force distribution in granular materials~\cite{liu1995force,coppersmith1996model,alim2017local}, it can be shown that the PDF of normalized fluid flux inside a highly disordered porous material converges to a universal distribution with an exponential form, i.e., $p(\hat{q}) \propto e^{-\alpha \hat{q}}$ for large $\hat{q}$ where $\hat{q}$ is the normalized fluid flux $\hat{q} = {q}/\langle q \rangle$ (see supplementary materials S3). This result indicates that the diamond grid network of pores where cylindrical edges between the pores have a highly disordered distribution is sufficient to capture the statistics of a heterogeneous fluid flow inside a porous material.

 % In the following parts we show the results of a diamond grid, and the results for a random-graph is presented in the supplementary material.

%  independent of the pore structure, becomes exponentially distributed. 
%  Considering a random \AZ{needs explicit definition} or structured network with a random distribution of tube's radius/length (i.e., effectively their flow resistance) and a 
%  For a large enough randomness in the edge resistances between the pores $R_{ij}$, the probability distribution of flux in the tubes normalized by the average flux follows a universal curve with an exponential tail, i.e., $P(\hat{q}) \propto e^{-\alpha \hat{q}}$ for large $\hat{q}$ where $\hat{q}$ is the normalized fluid flux $\hat{q} = {q}/\langle q \rangle$ (See Fig. \ref{fig:fig1}c and supplementary material for more details). This PDF of fluid flux $q$ has been shown to correctly captures the exponential tail for the fluid flux in a porous medium \cite{alim2017local,shima2021}.
%

>>>>

% ****** Start of file apssamp.tex ******
%
%   This file is part of the APS files in the REVTeX 4.1 distribution.
%   Version 4.1r of REVTeX, August 2010
%
%   Copyright (c) 2009, 2010 The American Physical Society.
%
%   See the REVTeX 4 README file for restrictions and more information.
%
% TeX'ing this file requires that you have AMS-LaTeX 2.0 installed
% as well as the rest of the prerequisites for REVTeX 4.1
%
% See the REVTeX 4 README file
% It also requires running BibTeX. The commands are as follows:
%
%  1)  latex apssamp.tex
%  2)  bibtex apssamp
%  3)  latex apssamp.tex
%  4)  latex apssamp.tex
%%
\documentclass[%
reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose,
% preprint,
%showpacs,preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
 % pra,
prl,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-1}

\usepackage{soul}
\usepackage{color}

\newcommand{\hlcyan}[1]{{\sethlcolor{cyan}\hl{#1}}}
\newcommand{\hlred}[1]{{\sethlcolor{red}\hl{#1}}}

@@

\usepackage{xcolor}
\newcommand{\DP}[1]{\noindent \color{cyan} (DP: #1)\normalcolor}
\newcommand{\AZ}[1]{\noindent \color{purple} (AZ: #1)\normalcolor}
\newcommand{\add}[1]{\noindent \color{blue} #1 \normalcolor}
\newcommand{\del}[1]{\noindent \color{red} \st{#1}\normalcolor}
\newcommand{\AM}[1]{\noindent \color{magenta} (AA: #1)\normalcolor}
\def\red{\textcolor{red}}
%
\def\blue{\textcolor{blue}}

\begin{document}
%
\preprint{APS/123-QED}

\title{Temporal Evolution of Erosion in Pore-Networks:\\ From Homogenization to Instability} 
% Force line breaks with \\
% \thanks{A footnote to the article title}%

\author{Ahmad Zareei}
\thanks{A.Z. and D.P. contributed equally.}
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break automatically or can be forced with 
\author{Deng Pan}%
\thanks{A.Z. and D.P. contributed equally.}
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break automatically or can be forced with
\author{Ariel Amir}%
\email{arielamir@seas.harvard.edu}
\affiliation{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02148}%\altaffiliation[]{SEAS, Pierce Hall}%Lines break  automatically or can be forced with
% need to figure out how to reference dagger

@@

\begin{abstract}
We study the dynamics of flow-networks in porous media using a two/three dimensional
pore-network model. We consider a class of erosion dynamics
for a single phase flow with no deposition, chemical reactions, or topology changes assuming a constitutive law depending on flow rate, local velocities, or shear stress at the walls. We show that depending on the erosion law, the flow may become uniform and homogenized or become unstable and develop channels. By defining an order parameter capturing these different behaviors we show that a phase transition occurs depending on the erosion dynamics. Using a simple model, we identify quantitative criteria to distinguish these regimes and {correctly predict the fate of the network, and discuss the experimental relevance of our result. }
%
% Lastly, we show that the initial stages of pores clogging show analogous behaviors which depends on the clogging dynamics.

% Lastly, we show that pores clogging at the initial stages show analogous behaviors depending on clogging dynamics. % , however, due to the pore throat blockages and changes in the network connections the evolution diverges from its initial trend.

@@

Fluid flow through a porous medium undergoing a dynamical change in its network of micro-structure is ubiquitous in nature~\cite{marbach2016pruning,alim2013random,tero2010rules,heaton2010growth} as well as in numerous environmental~\cite{schlesinger1999carbon,winkler1979pore,batzle1992seismic,cohen2015mechanisms} and industrial applications~\cite{duduta2011semi,sun2019hierarchical,smith2017multiphase,ferguson2012nonequilibrium}.
The disordered pore structure of a porous medium results in heterogeneously distributed fluid flow between the pores.
The boundaries of the pore structure can change dynamically either through erosion or deposition/sedimentation of material. Such heterogeneous changes of the solid structure affect the pore-level fluid flow which in turn affects the dynamical changes to the pore structure. This feedback mechanism along with the initial heterogeneous fluid flow makes it difficult to understand and predict the porous media behavior. Nonetheless, an understanding of the dynamical change is essential to improve any of the porous media applications where the pore network changes over time, including groundwater remediation and precipitation of minerals in rocks~\cite{rad2013pore}, biofilm growth in water filtration, and protective filters \cite{herzig1970flow,tien1979advances,jaisi2008transport,carrel2018biofilms,seymour2004anomalous}, as well as enhanced oil recovery with polymer flooding \cite{lake2014fundamentals,parsa2020origin}, or water-driven erosion \cite{schorghofer2004spontaneous,mahadevan2012flow}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
% \section{Methods \& Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
\begin{figure}[h]
    % \centering
    \includegraphics[width = 0.45\textwidth]{Fig1.pdf}
    % \caption{(a) Cross section of a porous sandstone sample obtained using  computerized tomography~\cite{akanji2010finite}. The scale in the bottom left shows $1$mm. The network of pores and throats is highlighted in blue. In the network model, the pores are represented with nodes and the throats between pores are approximated with tubes connecting the pores together. (b) Schematic of a topologically random network (i); and a structured diamond grid network (ii). The edge diameters in both networks are randomly distributed. The inset figure shows the conservation of mass at each node. (c) The universal probability distribution function (PDF) of fluid flux for a topologically random (in blue) or diamond network (red) with a highly disordered random network including uniform (triangles), log-normal (circles), and truncated normal (plus) distributions.}
    \caption{(a) Cross section of a porous sandstone sample obtained using  computerized tomography~\cite{akanji2010finite}. The scale in the bottom left shows $1$mm. The network of pores and throats is highlighted in blue. In the network model, the pores are represented with nodes and the throats between pores are approximated with tubes. (b) Schematic of a topologically random network. The edge diameters representing the pore-throats are randomly distributed. (c) A structured diamond grid network with a random distribution of edge diameters. (d) The universal probability density function (PDF) of fluid flux for a topologically random (blue) or diamond-grid (red) network of nodes (red) with a random distribution of edge diameter sampled from a uniform (triangles), log-normal (circles), or truncated normal (plus) distribution. }    
    \label{fig:fig1}
\end{figure}
\emph{Network approach}-- We approach this long-standing problem using a network model for the porous structure~\cite{fatt1956network, blunt2013pore,stoop2019disorder,bryant1993network,blunt2013pore,dong2009pore,blunt2013pore,blunt1995pore,alim2017local}.
% Before discussing the dynamics, we briefly review the results of flow in random networks.
The network of pores inside the solid structure is connected together through pore-throats that effectively show resistance to the fluid flow between the pores (Fig.~\ref{fig:fig1}a). Network-based models have been shown to successfully capture key properties of fluid flow in a porous material such as the probability distribution of fluid flux~\cite{alim2017local}, the permeability scaling during clogging~\cite{shima2021}, or the first fluidized path in a porous structure~\cite{fraggedakis_chaparian_tammisola_2021}.  We consider low-Reynolds fluid flow through the porous network. The fluid flow rate at the edge connecting pores $i$ and $j$ is given by $ q_{ij}  = C_{ij}(p_i - p_j)$ where $p_i,p_j$ represent pressures at neighboring nodes. Poiseuille's law implies that the conductance $C_{ij} = {\pi r_{ij}^4}/{8 \mu l_{ij}}$, with $r_{ij}$ and $l_{ij}$ the edge's radius and length. Initially, we consider a topologically random network of nodes constructed using uniformly distributed nodes (i.e., a Poisson point process) in a planar domain connected using Delaunay triangulation (Fig. \ref{fig:fig1}b). The pore-throats or radii of the edges are considered as independent and identically distributed random variables and fluid flow in the edges are obtained by solving for conservation of mass at all nodes given a pressure difference between the nodes on the boundaries (supplementary material, SM, \S1). Independent of edge radius distribution, the probability density function (PDF) of normalized fluid flux is well described by a single exponential distribution shown in Fig. \ref{fig:fig1}d. The exponential distribution of fluid flux is similar to earlier experimental and numerical measurements~\cite{datta2013spatial,shima2021,alim2017local} and is a universal feature in random porous networks. Considering a structured diamond-grid of pores (Fig. \ref{fig:fig1}c) which significantly simplifies the geometrical complexity of the network and allows for analytical tractability, one finds that the PDF of normalized fluid flux remains unchanged for various distributions suggesting robustness to network topology (Fig.~\ref{fig:fig1}d and SM \S2). In the following, we will study, analytically and numerically, how this universal distribution evolves as the network is modified based on a local constitutive law.

\textit{Network evolution model}-- The degradation of the solid skeleton (i.e., erosion) or deposition of material on the pore throats (i.e., clogging) in the network of pores is modeled by the change (increase or decrease) in the radii of the edges connecting the pores which translates into changes in the flow resistance between the pores. The rate of change of the radii depends on local fluid flow parameters (i.e., $q, r$), however, the exact dependence is unknown. A general constitutive model for erosion can be written as $dr_{ij}/dt = f(q_{ij},r_{ij})$, where $f(q_{ij},r_{ij})$ represents the functional dependence of erosion on the fluid flux $q_{ij}$ and the tube radius $r_{ij}$. Erosion in saturated, granular, porous medium occurs when the shear stress  at the walls $\tau_w$ exerted by the fluid overcomes the cohesive strength of the solid matter $\tau_c$. In a class of erosion models previously studied~\cite{jager2017channelization,ristroph2012sculpting,wan2004investigation} the eroded mass per unit area is linearly proportional to the excess shear, i.e.,  $\dot{m}  = -\kappa  (\tau_w-\tau_c)$ where $\kappa$ is the linear proportionality constant. This erosion model effectively results in an increase in the radius of the edges in our network model where \add{$\dot{r} \propto \kappa/r (|q|/r^3 - \tau_c)$ , where $1/r$ factor comes from the eroded mass $dm =2\pi\rho r l dr$ for a cylindrical pipe and $|q|/r^3$ comes from the known result of shear stress of laminar flow in pipes \cite{white2006viscous}}. Similar models have further been used in biological models for the growth of vascular networks~\cite{chen2012haemodynamics,hacking1996shear}. In addition to local shear stress, erosion based on local fluid velocity ($v \propto |q|/r^2$) \cite{kudrolli2016evolution}, local pressure difference ($\Delta p \propto |q|/r^4$) \cite{derr2020flow,mahadevan2012flow}, and power dissipation ($P \propto q^2/r^4$ ) ~\cite{steeb2007modeling,marot2012study,sibille2015internal} have also been considered as alternative models.  Here, we first focus on a class of erosion dynamics described by $f(q_{ij},r_{ij}) = \alpha |q_{ij}|^m/r_{ij}^n$ where $m,n,\alpha>0$ are constants. 
This model unifies a diverse set of erosion and clogging dynamics as different values of $m$ and $n$ correspond to different physics of erosion. Later we will generalize the models to include other forms of nonlinearities. % We use a general constitutive  model which may implement a diverse set of erosion or clogging dynamics and thus allows us to study the effect of different laws in a unified way. 
\begin{figure}[!h]
    % \centering
    % \includegraphics[width = 0.48\textwidth]{Figs/Fig_2_Random_Network.png}
    \includegraphics[width = 0.48\textwidth]{Fig2_.pdf}
    \caption{Erosion in a network of pipes. The initial condition is shown with the label $t=0$ in the first row. Each row afterward corresponds to the {simulation result at $t=T$ where  $\langle r_{t=T}\rangle=2r_0$ with $r_0 = \langle r_{t=0}\rangle$}. The erosion law is based on $f(|q|,r) = |q|/r^n$  with different powers of $n$ correspond to different models of erosion. The first, second, and third columns are snapshots of the pore network, the PDF of normalized fluid flux $q/\langle q \rangle$, and  the PDF of normalized radius $r/\langle r\rangle$ at $t=T$, respectively.} \label{fig:fig2}
\end{figure}
% \begin{figure*}[htp]
%      % \centering
%       \includegraphics[width = 1.0\textwidth]{FigS2.pdf}
%      \caption{\add{Erosion in a topologically random 3D network of tubes with $N_x=20, N_y=12, N_z=12$ points for Voronoi cell initialization and an initial uniform broad distribution of tube diameters randomly sampled from $\mathcal{U}(1,14)$. Snapshots of the network, PDF of normalized fluid flux $q/\langle q \rangle$, and normalized edge radius distribution $r/\langle r \rangle$ at the initial time $t=0$, and also after $N$ erosion steps for different powers of erosion $n$ are shown. We stop the erosion after $N$ steps such that $\langle r\rangle=2r_0$ where $r_0 = \langle r_{t=0}\rangle$. The erosion law is based on Eq. (1) \del{in the main text} where different powers of $n$ correspond to different models of erosion.}}\label{SIfig:fig2-3d}
%  \end{figure*}
% In order to model erosion in porous media, we consider the abrasion in the throats leading to decrease in the tube radii. We model the dynamics as 
% \begin{align}
%   \frac{dr_{ij}}{dt} = \alpha  \frac{|q_{ij}|^m}{r_{ij}^n}, \label{eq:erosion-law}
% \end{align}
% where $m,n,\alpha$ are constants. Different values of $m$ and $n$ along with a positive $\alpha>0$ correspond to different erosion physics (analogously, $\alpha<0$ corresponds to clogging).
Particularly, the erosion when $m=1$ and  (i) $n=0$ depends on the amount of flux $q_{ij}$ passing through the edge; (ii) $n=2$ depends on the local velocities  \cite{kudrolli2016evolution}; (iii) $n=3$ depends on the shear force at the boundary of the throat~ \cite{bonelli2011micromechanical,parker2000purely,jager2017channelization}. {Additionally, $m=2$ and $n=6$ corresponds to models considered in biological transport networks where the radii changes are {proportional to the square of shear stress at the boundary walls}~\cite{hu2013adaptation,ronellenfitsch2016global}.}  We later address the effect of incorporating the threshold discussed above into our model.  
We consider a randomly initialized network with disordered diameters obtained from a uniform distribution (SM S1). 
The flow inside the pores, PDF of flux in the tubes, and PDF of tube radii are shown in Fig. \ref{fig:fig2}. We assume a constant pressure difference between the left and the right boundaries. In each time step, we increase the local radii of the tubes based on the erosion law $dr_{ij}/dt = \alpha q_{ij}^m/r_{ij}^n$, assuming erosion is linear in the flux ($m=1$). Later we will consider the network behavior for other powers of $m$. 
We continue the simulations until $\langle r \rangle = 2 r_0$. 
The results of the simulations for different values of $n$ are shown in Fig. \ref{fig:fig2}. When $n=1$ or $2$,  the network develops channels. In such cases, the flow is dominated by a few edges carrying most of the flow while the rest of the network carries almost no flow. This is also reflected by a bimodal radii distribution. In contrast, when $n=3$  (corresponding to erosion linear in shear), we find that despite the 
increase in tube radii and absolute flow rates, the normalized flow distribution is hardly affected, maintaining its original exponential form. Increasing $n$ to larger values, $n=4$ or $5$, we find that the flow pattern in the network moves towards homogenization. Here, the tail of the normalized fluid flux distribution retracts and the coefficient of variation reduces. Similarly, the PDF of the tube diameters becomes narrower.  It is to be noted that the different distributions obtained for different powers of $m,n$ represent different statistics that cannot be collapsed.

We found that similar results uphold in a 3d random tube network  (see SM, Fig.~S2) as well as a 2d topologically ordered (diamond grid) network, illustrating the robustness of the results to the network topology (SM, Fig.~S3). Similarly, we found that the results hold also when using an initial narrow distribution of diameters (SM, Fig.~S4), showing robustness with respect to the strength of the disorder.  We further changed the boundary conditions, using a circular random network with the inlet at the center and outlet at the outer edges and we find that similar results uphold~(SM, Fig.~S5).

\textit{Phase transition and order parameter}-- {To quantify the transition of the network between the channeling instability and homogenization, we define an order parameter % $\mathcal{O}$ 
%
\begin{align}
\mathcal{O} = \frac{1}{N-1} \left( N - \frac{\left( \sum_{ij} q^2_{ij}\right)^2}{ \sum_{ij} q^4_{ij}}\right),
\end{align}
%
where $N$ is the number of edges.} {The order parameter defined here is inspired by the participation ratio (PR) employed to quantify the localization of an eigenstate in the analysis of Anderson localization~\cite{kramer1993localization}.}  The order parameter $\mathcal{O}=0 $ when the flux through every edge is identical. On the other hand, when fluid flux becomes highly localized with only a few edges with non-zero flux, $\mathcal{O}\to 1$.  We numerically calculated the  order parameter $\mathcal{O}$ for {randomly initialized networks, averaged over 20 different realization}. 
The results are shown in Fig. \ref{fig:fig3}a for different amounts of erosion measured by the increase in the average diameter $\langle r\rangle /r_0$. As shown in 
 Fig. \ref{fig:fig3}c, at $n\approx 3$ the order parameter, remains unchanged; however for $n>3$ the order parameter moves toward zero, where the flow becomes more uniform, and for $n<3$ the order parameter goes toward unity, where channels are developed. This indicates a phase transition at $n=3$ in the long-time behavior of the network.
%
